---
title: "R for Everyone Part 5: Loops"
description: "Learning Objective: Create loops to add iterative flows."
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)
```


## Getting Started

Take a look at the code below. Try to determine the resulting value of `x` without running the code.

```{r exercise_03a, exercise=TRUE}
x <- 1

if (x < 7){
  x <- x + 1
}

if (x < 7){
  x <- x + 1
}

if (x < 7){
  x <- x + 1
}

if (x < 7){
  x <- x + 1
}

if (x < 7){
  x <- x + 1
}

if (x < 7){
  x <- x + 1
}

if (x < 7){
  x <- x + 1
}
```

```{r exercise_03b, echo=FALSE}
question("What is the value of x?",
  answer("6"),
  answer("7", correct = TRUE),
  answer("8"),
  answer("Something else"),
  allow_retry = TRUE
)

```

That might've been an annoying exercise, and you had to count the number of times `if (x < 7)` appeared in the code to ensure the answer wasn't 6.

Luckily, we can use loops to simplify this code and make it more human-interpretable! **Loops** are used to express code that repeats or iterates. In the sample above, you saw that `if (x < 7)` repeated many times.

There are two common forms of loops in R. The `while` loop and the `for` loop.

## While Loop

The **while loop** takes the form below:

    while (boolean_expression){

      # Execute some_code if `boolean_expression` is `TRUE`
      some_code 
      
    }

This may look very similar to an if statement, but there is a key difference. After executing `some_code`, the while loop will check the `boolean_expression` again, and if it is `TRUE`, then the loop will continue executing `some_code`. The while loop will keep repeating or looping until the `boolean_expression` is `FALSE`.

Using a while loop, we can dramatically simplify our code:

```{r exercise_03c, exercise=TRUE}
x <- 1

while (x < 7){
  x <- x + 1
}

print(x)
```

A common error involving while loops is when the `boolean_expression` never evaluates to FALSE, and the loop continues looping indefinitely. This is called an **infinite loop**. Below is an example of an infinite loop.

    x <- 1

    # This will infinite loop because x is always greater than 0!
    while (x > 0){ 
      x <- x + 1
    }

***Pro Tip**: If you are stuck in an infinite loop while executing R code, don't panic. To halt the infinite looping, click the stop sign icon that appears in the top right of the Console pane in RStudio.*

## For Loop

The **for loop** is slightly different than a while loop. Rather than using a condition to determine whether to continue looping, the for loop will apply some code to every item in a vector. For example:

```{r exercise_03d, exercise=TRUE}
# For every `int_value` in `1:7`, assign `int_value` to `x`
for (int_value in 1:7){ 
  x <- int_value
}

print(x)
```

*Note: You haven't learned about vectors yet. For now, think of them as collections of items*

## Knowledge Check

Write some code based on the description below:

    Assign 0 to an object named `int_value`
    Increment `int_value` by 3 while it is less than or equal to 999
    Print the value stored in `int_value`

You can reveal the solution once you are comfortable with your response

```{r KC3, exercise=TRUE, exercise.lines=15}
# YOUR WORK HERE
# Assign 0 to an object named `int_value`


# Increment `int_value` by 3 while it is less than or equal to 999




# Print the value stored in `int_value`

```

```{r KC3-solution}
# Assign 0 to an object named `int_value`
int_value <- 0

# Increment `int_value` by 3 while it is less than or equal to 999
while (int_value <= 999){
  int_value <- int_value + 3
}

# Print the value stored in `int_value`
print(int_value)
```

```{r KC3-question, echo=FALSE}
question("What is the end value of `int_value`?",
  answer("999"),
  answer("1000"),
  answer("1001"),
  answer("1002", correct = TRUE),
  allow_retry = TRUE
)
```

## Summary

| Vocabulary                    | Definition                                                                                           |
|:------------------------------|:-----------------------------------------------------------------------------------------------------|
| Loops                         | Code that repeats or iterates.                                                                       |
| While Loop                    | A type of loop that iterates based on a condition.                                                   |
| For Loop                      | A type of loop that iterates through a vector.                                                       |
