---
title: "R for Everyone Part 4: Conditions"
description: "Learning Objective: Apply conditional logic using if...else statement."
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)
```


## Getting Started

Up to this point, the computer programs (R scripts) you created all executed sequentially. In other words, R ran every line of your code in order for top to bottom.

In many cases, you may want your code to run differently, and you need to apply **control structures** to control the flow of your computer program. The first type of control structure you'll learn about is the condition.

## What is a Condition?

In real life, conditions indicate situations or circumstances that must be true in order for something to happen. Below are some examples of statements with conditions (conditions are *italicized*):

-   "Make sure to drip your faucets *when the temperature drops below 15 degrees Fahrenheit*."

-   "We cannot run a marathon unless *we prepare and train regularly*."

-   "*If you have a question*, please raise your hand."

In computer programming, **conditions** indicate something that must be true in order for some code to be executed. Notice how the phrase *must be true* appears in our definition of conditions in both real life and computer programming. Let's dig a little deeper and see how we determine whether something is true in computer programming.

## Boolean Expressions

To determine whether something is true we use boolean expressions. **Boolean** expressions result in boolean values, and boolean values are values that are either `TRUE` or `FALSE` . Similar to other types of values, boolean values can be stored inside of objects. Let's take a look at some boolean values and expressions below.

```{r exercise_02a, exercise=TRUE}
# Assign to an object called `x`, the result of `1 == 1`
x <- 1 == 1

# Assign to an object called `y`, the result of `2 < 1`
y <- 2 < 1
```

```{r exercise_02b, echo=FALSE}
quiz(
  caption = "Answer the following questions:",
  question("What is the value of x? *Hint: Print the value of x in the code block above.*",
    answer("TRUE ", correct = TRUE),
    answer("FALSE "),
    answer("1"),
    answer("2"),
    allow_retry = TRUE
  ),
  question("What is the value of y? *Hint: Print the value of y in the code block above.*",
    answer("TRUE "),
    answer("FALSE ", correct = TRUE),
    answer("1"),
    answer("2"),
    allow_retry = TRUE
  )
)

```

The `==` and `<` operators are just two examples of operators that produce boolean values. There are others, and they are listed below:

| Operator | Description                                                                                       | Example                       |
|----------|---------------------------------------------------------------------------------------------------|-------------------------------|
| `<`      | **Less than:** TRUE if the left value is less than the right value.                               | `1 < 2 #This is TRUE`         |
| `<=`     | **Less than or equal to:** TRUE if the left value is less than or equal to the right value.       | `1 <= 1 #This is FALSE`       |
| `>`      | **Greater than:** TRUE if the left value is greater than the right value.                         | `1 > 1 #This is FALSE`        |
| `>=`     | **Greater than or equal to:** TRUE if the left value is greater than or equal to the right value. | `3 >= 2 #This is TRUE`        |
| `==`     | **Equal to:** TRUE if the left value is equal to the right value.                                 | `3 == 2 #This is FALSE`       |
| `!=`     | **Not equal to:** TRUE if the left value is not equal to the right value.                         | `1 != 1 #This is FALSE`       |
| `!`      | **Not:** TRUE if the value is FALSE, FALSE if the value is TRUE.                                  | `! TRUE #This is FALSE`       |
| `|`      | **Or:** TRUE if either the left value or right value are TRUE.                                    | `TRUE | FALSE #This is TRUE`  |
| `&`      | **And:** TRUE if both the left value and right value are TRUE.                                    | `TRUE & FALSE #This is FALSE` |

Similar to other operators, these operators can be used with variables (e.g. `x <= y`).

## If...Else Statements

Now that you know about boolean expressions, you're ready to learn about conditions in R. Specifically, you'll learn about **if...else** statements.

The most basic form of an if...else statement executes some code if a boolean expression is TRUE. To do this, you'll need to use `if`. Take a look below:

```{r exercise_02c, exercise=TRUE}
# Assign to an object named `x` the value `1`
x <- 1

if (x > 0){     # Check whether `x` is greater than `0`
  x <- x + 1    # if `x > 0` is TRUE, then add `1` to `x`
}
```

```{r exercise_02c_mc, echo=FALSE}
question("What is the value of x?",
  answer("1"),
  answer("2", correct = TRUE),
  answer("TRUE"),
  answer("FALSE"),
  allow_retry = TRUE
)
```

Sometimes, you may want to do something else if the boolean expression turns out to be FALSE. To do this, you'll need to include an `else`. Take a look below:

```{r exercise_02d, exercise=TRUE, exercise.diagnostics=FALSE}
# Assign to an object named `y` the value `10`
y <- 10

if (y < 5){     # Check whether `y` is less than `5`
  y <- y * 50   # If `y < 5` is TRUE, multiply `y` by `50`
} else {        # Otherwise...
  y <- 1        # Assign `1` to `y`
}
```

```{r exercise_02d_mc, echo=FALSE}
question("What is the value of y?",
  answer("10"),
  answer("500"),
  answer("5"),
  answer("1", correct = TRUE),
  allow_retry = TRUE
)
```

In other cases, you may want to check multiple conditions and perform different actions depending on which is TRUE. To do this, you'll need to use `else if`. Take a look below:

```{r exercise_02e, exercise=TRUE, exercise.diagnostics=FALSE}
# Assign to an object named `z` the value `50`
z <- 50

if (z > 5){               # Check whether `z` is greater than `5`
  z <- z * 50             # If `z > 5` is TRUE, then multiply `z` by `50`
} else if (z > 200){      # Otherwise, check if `z > 200`
  z <- 50                 # If `z > 200` is TRUE, Assign `50` to `z`
} else {                  # Otherwise...
  z <- 1                  # If none of the conditions above are TRUE, assign `1` to `z`
}
```

```{r exercise_02e_mc, echo=FALSE}
question("What is the value of z?",
  answer("50"),
  answer("2500", correct = TRUE),
  answer("5"),
  answer("1"),
  allow_retry = TRUE
)
```

You can have multiple `else if`s chained together if you have many conditions to evaluate.

Generalizing what you saw in the above scenarios, conditional statements in R take the **if...else** form below:

    if (boolean_expression){

      # Execute some_code if `boolean_expression` is `TRUE`
      some_code 
      
    } else if (another_boolean_expression){

      # Execute `some_other_code` if `boolean_expression` is `FALSE` and `another_boolean_expression` is `TRUE`
      some_other_code 
      
    } else {

      # Execute `some_more_code` if all the above conditions are `FALSE`
      some_more_code 
      
    }

## Knowledge Check

Write some code based on the description below:

    Subtract 85 from 100 and assign the result to an object named `score`

    If score is 90 or above, assign 'A' to an object named `grade`

    Otherwise if the `score` is 80 or above, assign 'B' to an object named `grade`

    Otherwise if the `score` is 70 or above, assign 'C' to an object named `grade`

    Otherwise if the `score` is 60 or above, assign 'D' to an object named `grade`

    Otherwise if the `score` is below 60, assign 'Hidden' to an object named `grade`

    Print the value stored in `grade`

You can reveal the solution once you are comfortable with your response

```{r KC2, exercise=TRUE, exercise.lines=15, exercise.diagnostics=FALSE}
# YOUR WORK HERE
# Subtract 85 from 100 and assign the result to an object named `score`

# If score is 90 or above, assign "A" to an object named `grade`

# Otherwise if the `score` is 80 or above, assign "B" to an object named `grade`

# Otherwise if the `score` is 70 or above, assign "C" to an object named `grade`

# Otherwise if the `score` is 60 or above, assign "D" to an object named `grade`

# Otherwise if the `score` is below 60, assign "Hidden" to an object named `grade`

# Print the value stored in `grade`

```

```{r KC2-solution}
# Subtract 85 from 100 and assign the result to an object named `score`
score = 100 - 85

if (score >= 90){          # If score is 90 or above, assign "A" to an object named `grade`
  grade <- "A"
} else if (score >= 80){   # Otherwise if the `score` is 80 or above, assign "B" to an object named `grade`
  grade <- "B"
} else if (score >= 70){   # Otherwise if the `score` is 70 or above, assign "C" to an object named `grade`
  grade <- "C"
} else if (score >= 60){   # Otherwise if the `score` is 60 or above, assign "D" to an object named `grade`
  grade <- "D"
} else{                   # Otherwise if the `score` is below 60, assign "Hidden" to an object named `grade`
  grade <- "Hidden"
}

# Print the value stored in `grade`
print(grade)
```

## Summary

| Vocabulary                    | Definition                                                                                           |
|:------------------------------|:-----------------------------------------------------------------------------------------------------|
| Control Structures            | Control the order in which R runs your code.                                                         |
| Condition                     | Indicates something that must be true in order for some code to be executed.                         |
| Boolean                       | Indicates logical TRUE or FALSE.                                                                     |
| If...Else Statements          | A control structure in R that runs code base on conditions.                                          |
